Mathematical Aspects of the Capital Market

### 2017-08-09 phd/Chapterintroduction.tex
Added texts to the introduction section of the phd thesis.

Kesten \cite{kesten:1973} showed that, when $A_t$ are almost
surely non-negative, has no row or column of only zeros, and
$B_t$ is almost surely non-negative and is not equal to the null
vector with probability 1, then the solution $X$ to the equation
$X \overset{d}{=} A X + B$ is regularly varying with a positive index
$\alpha$, assuming the following conditions (M) and (A):
\begin{itemize}
\item Condition (M)
  \begin{enumerate}
  \item The top Lyapunov exponent
    \[
    \gamma = \inf_{n \geq 1} {1 \over n}\E \log \|A_n \cdots A_1\|
    \]
    is negative.
  \item There exists $\alpha > 0$ such that
    \[
    1 = \lambda(\alpha) = \lim_{n \to \infty} {1 \over n} \log \E \|A_n \cdots A_1\|^\alpha
    \]
  \item $\E (\|A_1\|^\alpha \log^+\|A_1\|) < \infty$
  \item $\E |B_1|^\alpha < \infty$
  \end{enumerate}
\item Condition (A) : The group generated by
  \[
  \{\log\rho(s): s = A_n \cdots A_1 \text{ for some } n \geq 1\}
  \]
  is dense in $\reals$, where $\rho(s)$ denotes the spectral
  radius of matrix $s$.
\end{itemize}
Upon these conditions, Kesten's theorem gives
\begin{equation*}
  u^\alpha \P(u^{-1} V \in \cdot) \overset{v}{\to} \nu_\alpha(\cdot)
\end{equation*}
where $\nu_\alpha$ is a non-null Radon measure on
$\reals^d_+ \setminus \{0\}$ with the property
$\nu_\alpha(a A) = a^{-\alpha} \nu_\alpha(A)$.

\section{GARCH models}
Introduced by Bollerslev \cite{bollerslev:1986} in 1986,
{\em Generalized Autoregressive Conditional Heteroscedasticity}
(GARCH) models have been hugely popular for modelling volatility of
financial time series and have inspired numerous variants.
A GARCH($p, q$) model is a stationary process
$\{X_t\}_{t \in \mathbb Z}$satisfying
\begin{eqnarray*}
  X_t &=& \sigma_t Z_t \\
  \sigma_t^2 &=& \omega + \sum_{i=1}^p \alpha_i X_{t-i}^2 +
  \sum_{j=1}^q \beta_j \sigma_{t-j}^2
\end{eqnarray*}
where $\{X_t\}$ is a return series, e.g. stock returns, foreign exchange
rates, interest rates, etc; $\{Z_t\}$ is a iid, mean 0, unit-variance
sequence; and $\sigma_t^2$ is the variance of the
distribution of $X_t$ conditional on $\{(X_i,
\sigma_i^2)\}_{i=0}^{t-1}$; $\omega, \{\alpha_i\}_{i=1}^p,
\{\beta_i\}_{i=1}^q$ are non-negative parameters of the
model. The GARCH($p,q$) recurrence equation is of the form of
\eqref{eq:rhjyu}. With appropriate conditions, $\sigma_t^2$ is
shown to be a positive Harris recurrent Markov chain (cf. Bollerslev
\cite{bollerslev:1986} and Buraczewski et al.
\cite{buraczewski:damek:mikosch:2016}), whose stationary distribution
has regularly varying tails. The tail index $\alpha$, is given by
\begin{equation}
  \label{eq:grgbg}
  \lim_{n \to \infty} {1 \over n}\log\E\|A_n \cdots A_1\|^\alpha = 0,
\end{equation}
where $\{A_i\}_{i \in \mathbb Z}$ are iid matrices whose entries are
functions of $\{\alpha_i\}_{i=1}^p$, $\{\beta_i\}_{i=1}^q$ and
$\{Z_t^2\}$:
\[
A_i =
\begin{pmatrix}
  \alpha_1 Z_{t-1}^2 + \beta_1 & \beta_2 & \cdots &
  \beta_{q-1} & \beta_q & \alpha_2 & \alpha_3 &
  \cdots & \alpha_{p-1} & \alpha_p\\
  1 & 0 & \cdots & 
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots & 
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  1 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  Z_{t-1}^2 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  0 & 0 & 1 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots &
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\    
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 1 & 0 \\    
\end{pmatrix}
\]
While GARCH models have been very successful for modelling financial
time series, it does have its drawbacks. For example, the tail
index is very sensitive to the model parameters $\{\alpha_i\}_{i=1}^p$ and
$\{\beta_i\}_{i=1}^q$. In applications, these parameters need to be
estimated from a sample and are all ways uncertain to an extent, there
can be a significant discrepancy between the tail index implied by
\eqref{eq:grgbg} and the Hill estimate \eqref{eq:fhhty}.

@@

we see that we are again in the framework of a stochastic recurrence equation\ 
but this time for vector-valued $\bfB_t$ and matrix-valued $\bfA_t$:

@@

\section{Stochastic Volatility models}

>>>>

Kesten \cite{kesten:1973} showed that, when $A_t$ are almost
surely non-negative, has no row or column of only zeros, and
$B_t$ is almost surely non-negative and is not equal to the null
vector with probability 1, then the solution $X$ to the equation
$X \overset{d}{=} A X + B$ is regularly varying with a positive index
$\alpha$, assuming the following conditions (M) and (A):
\begin{itemize}
\item Condition (M)
  \begin{enumerate}
  \item The top Lyapunov exponent
    \[
    \gamma = \inf_{n \geq 1} {1 \over n}\E \log \|A_n \cdots A_1\|
    \]
    is negative.
  \item There exists $\alpha > 0$ such that
    \[
    1 = \lambda(\alpha) = \lim_{n \to \infty} {1 \over n} \log \E \|A_n \cdots A_1\|^\alpha
    \]
  \item $\E (\|A_1\|^\alpha \log^+\|A_1\|) < \infty$
  \item $\E |B_1|^\alpha < \infty$
  \end{enumerate}
\item Condition (A) : The group generated by
  \[
  \{\log\rho(s): s = A_n \cdots A_1 \text{ for some } n \geq 1\}
  \]
  is dense in $\reals$, where $\rho(s)$ denotes the spectral
  radius of matrix $s$.
\end{itemize}
Upon these conditions, Kesten's theorem gives
\begin{equation}
  \label{eq:jorg}
  u^\alpha \P(u^{-1} V \in \cdot) \overset{v}{\to} \nu_\alpha(\cdot)
\end{equation}
where $\nu_\alpha$ is a non-null Radon measure on
$\reals^d_+ \setminus \{0\}$ with the property
$\nu_\alpha(a A) = a^{-\alpha} \nu_\alpha(A)$.

In addition to non-negative matrices, two other classes of random
matrices have been shown to lead to power-law tails via the recurrence
relation \eqref{eq:garchpq_sre}. Alsmeyer and Mentemeier
\cite{alsmeyer:mentemeier:2012} considered invertible matrices whose
distribution has a density. Let $M(d, \reals)$ denote the space of
$d \times d$ matrices with real entries that are invertible with
probability 1. They replaced Kesten's condition
$\E (\|A\|^\xi \log^+\|A\|) < \infty$ with a stronger
counterpart
$\E [\|A\|^\xi (\log^+\|A\| + \log \|A^{-1} \|)] < \infty$,
and lifted the condition (A). In addition, they assumed
\begin{enumerate}
  \item The Markov chain $X_n$ on $\sphere^{d-1}$, namely
    $X_n = A_n X_{n-1} / |A_n X_{n-1}|$, is irreducible, i.e. for any open
    set $U \subset \sphere^{d-1}$ and any $u \in \sphere^{d-1}$, 
    $\exists n \geq 1$ such that $\P(X_n u \in U) > 0$.
  \item There exist $N \geq 1$, $c, \epsilon > 0$ and an invertible
    matrix $\bar A \in M(d, \reals)$ such that for any set
    $C \subset M(d, \reals)$, it holds true
    $\P(A_N \cdots A_1 \in C) \geq c |B_\epsilon(\bar A) \cap C|$,
    where $|\cdot|$ denotes the Lebesgue measure.
\end{enumerate}
These assumptions are termed conditions (id). Furthermore, they assumed
that there was no point in $\reals^d$ such that the recurrence
equation \eqref{eq:garchpq_sre} was stuck at this point with probability 1: 
$\P(A X + B = X) < 1$ for all $X \in \reals^d$ and all $A \in M(d, \reals)$.
With these assumptions, they showed
\[
\lim_{u \to \infty} u^\xi \P(\inn{x, V} > u) = e_\xi(x)
\]
where $x \in \sphere^{d - 1}$ and $e_\xi(\cdot)$ is a continuous
function $\sphere^{d-1} \to \reals_+$.

The second of the (id) conditions, which is satisfied when the
distribution of $A$ has a Lebesgue density, can actually be lifted if
stronger moment conditions are imposed on $A$ and $B$, and in
addition, a proximity condition is satisfied by the support of
$A$. This is the result of Guivarc'h and Le Page, et al.
\cite{guivarc:page:2016}. Let $G_A$ denote the semi-group generated
by $\{\Pi_n: \Pi_n = A_n \cdots A_1, A_i \in M(d, \reals)\}$. The
authors assumed
\begin{enumerate}
  \item There is no finite union $W$ of proper subspaces of $\reals^d$
    that satisfies $\forall a \in G_A, a W = W$.
  \item $G_A$ contains a proximal element, i.e. an element $a$ whose
    largest singular value is an algebraically simple eigenvalue of $a$.
\end{enumerate}
These two assumptions are termed (ip) conditions. Replacing the (id)
conditions of Alsmeyer and Mentemeier with (ip) and the moment
conditions of the former with
\[
\E \|A\|^{\xi + \delta} < \infty, \quad
\E (\|A\|^{\xi} \|A^{-1}\|^{\delta}) < \infty, \quad
\E (|B|^{\xi + \delta} < \infty),\quad
\text{ for some }\delta > 0
\]
Guivarc'h and Le Page et al. proved the same vague convergence result
of \eqref{eq:jorg}.

In the special case of GARCH($p, q$) processes, which are discussed in
more details in the next section, Bollerslev \cite{bollerslev:1986}
showed that the fixed-point equation $X \overset{d}{=} A X + B$ has a
unique, strictly stationary solution $X$ with finite variance if and
only if
\begin{equation}
  \sum_{i=1}^p \alpha_i + \sum_{j=1}^q \beta_j < 1
  \label{eq:bollerslev:intro}
\end{equation}

\section{GARCH models}
Introduced by Bollerslev \cite{bollerslev:1986} in 1986,
{\em Generalized Autoregressive Conditional Heteroscedasticity}
(GARCH) models have been hugely popular for modelling volatility of
financial time series and have inspired numerous variants.
A GARCH($p, q$) model is a stationary process
$\{X_t\}_{t \in \mathbb Z}$satisfying
\begin{eqnarray*}
  X_t &=& \sigma_t Z_t \\
  \sigma_t^2 &=& \omega + \sum_{i=1}^p \alpha_i X_{t-i}^2 +
  \sum_{j=1}^q \beta_j \sigma_{t-j}^2
\end{eqnarray*}
where $\{X_t\}$ is a return series, e.g. stock returns, foreign exchange
rates, interest rates, etc; $\{Z_t\}$ is a iid, mean 0, unit-variance
sequence; and $\sigma_t^2$ is the variance of the
distribution of $X_t$ conditional on $\{(X_i,
\sigma_i^2)\}_{i=0}^{t-1}$; $\omega, \{\alpha_i\}_{i=1}^p,
\{\beta_i\}_{i=1}^q$ are non-negative parameters of the
model. The GARCH($p,q$) recurrence equation is of the form of
\eqref{eq:rhjyu}. With appropriate conditions, $\sigma_t^2$ is
shown to be a positive Harris recurrent Markov chain (cf. Bollerslev
\cite{bollerslev:1986} and Buraczewski et al.
\cite{buraczewski:damek:mikosch:2016}), whose stationary distribution
has regularly varying tails. The tail index $\alpha$, is given by
\begin{equation}
  \label{eq:grgbg}
  \lim_{n \to \infty} {1 \over n}\log\E\|A_n \cdots A_1\|^\alpha = 0,
\end{equation}
where $\{A_i\}_{i \in \mathbb Z}$ are iid matrices whose entries are
functions of $\{\alpha_i\}_{i=1}^p$, $\{\beta_i\}_{i=1}^q$ and
$\{Z_t^2\}$:
\[
A_i =
\begin{pmatrix}
  \alpha_1 Z_{t-1}^2 + \beta_1 & \beta_2 & \cdots &
  \beta_{q-1} & \beta_q & \alpha_2 & \alpha_3 &
  \cdots & \alpha_{p-1} & \alpha_p\\
  1 & 0 & \cdots & 
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots & 
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  1 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  Z_{t-1}^2 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  0 & 0 & 1 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots &
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\    
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 1 & 0 \\    
\end{pmatrix}
\]
While GARCH models have been very successful for modelling financial
time series, it does have its drawbacks. For example, the tail
index is very sensitive to the model parameters $\{\alpha_i\}_{i=1}^p$ and
$\{\beta_i\}_{i=1}^q$. In applications, these parameters need to be
estimated from a sample and are all ways uncertain to some extent. For
this reason, there can be a significant discrepancy between the tail
index obtained via \eqref{eq:grgbg} with the model parameters
substituted for their sample estimates and the Hill estimate
\eqref{eq:fhhty}.

@@

we see that we are again in the framework of a stochastic recurrence
equation but this time for vector-valued $\bfB_t$ and matrix-valued
$\bfA_t$:

@@

\section{Stochastic Volatility models}
With the availability of high-frequency data, a different approach
than that of GARCH has been popularized for modelling volatility of
financial time series and has led to greatly improved accuracy of
prediction. This is the approach of stochastic volatility models.

In the pioneering work of Clark \cite{clark:1973}, the author modelled
the logarithmic price $X_t$ as a subordinated stochastic process:
$X_t = V_{\tau_t}$, where $V_i$ is a normally distributed random
variable with mean 0, variance $\sigma^2 i$ and independent
increments. $\tau_t$ is a real-valued, non-negative, non-decreasing
sequence with $\tau_0 = 0$. It models a time change. As pointed out by
Shepard and Andersen \cite{Shephard:Andersen:2009}, the log-price
process $X_t$ is serially uncorrelated although potentially dependent,
provided that $V_t$ and $\tau_t$ are independent.

Later authors, e.g. Back \cite{back:1991} chose to model the log-price
process as a semimartingale, with increments of the Martingale
component modelled as a product process:
\begin{eqnarray*}
  Y_t &=& Y_0 + A_t + M_t \\
  M_t - M_{t-1} &=& X_t = \sigma_t Z_t
\end{eqnarray*}
where $M_t$ is a semimartingale, $A_t$ is a finite variation process
and $M_t$ is a Martingale. $\sigma_t$ is non-negative and $Z_t$ is an
iid process with 0 mean and unit variance. A convenient choice of
$\sigma_t$ is
\[
\log\sigma_t = \sum_{l \in \mathbb Z} \psi_l \eta_{t-l}
\]
where $\psi_l$ is a sequence of non-negative numbers with at least one
positive element and $0 \leq \psi_l \leq 1$.

Clearly, if either
